---
title: "FINAL — Cleaned B-lineage patient-level disease comparison"
subtitle: "Exploratory analysis approved after manual review"
format:
  html:
    embed-resources: true
    self-contained-math: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: false
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(data.table)
library(jsonlite)
library(knitr)

result_dir <- Sys.getenv("BLINEAGE_DISEASE_RESULT_DIR")
if (!nzchar(result_dir)) stop("BLINEAGE_DISEASE_RESULT_DIR is not set")
result_dir <- normalizePath(result_dir, winslash = "/", mustWork = TRUE)
table_dir <- file.path(result_dir, "tables")
figure_dir <- file.path(result_dir, "figures")

read_table <- function(name) fread(file.path(table_dir, name))
primary <- read_table("07_primary15_model_results_with_global_bh_fdr.tsv")
sensitivity <- read_table("09_j2_l3_only13_sensitivity_model_results.tsv")
stability <- read_table("11_primary15_vs_j2_l3_only13_direction_stability.tsv")
panel_stability <- read_table("17_panel_level_direction_consistency_summary.tsv")
group_stats <- read_table("06_group_descriptive_statistics.tsv")
patients <- read_table("05_patient_level_raw_metrics.tsv")
predictions <- read_table("10_adjusted_group_predictions.tsv")
diagnostics <- read_table("12_model_diagnostics.tsv")
influence <- read_table("13_patient_influence_diagnostics.tsv")
loo <- read_table("15_leave_one_patient_out_summary.tsv")
direction <- read_table("16_same_bin_rook_cpm_descriptive_direction_consistency.tsv")
validation <- read_table("19_output_validation.tsv")

fdr_n <- primary[bh_fdr_18 < 0.05, .N]
ci_cross_zero_n <- primary[ci_low <= 0 & ci_high >= 0, .N]
all_consistent <- panel_stability[all_three_direction_consistent == TRUE, panel]
flipped <- stability[stability_class == "direction_flipped"]
ssc_ipf_primary_all_negative <- primary[
  contrast == "SSc-ILD vs IPF",
  all(effect_logit < 0)
]
ssc_ipf_sensitivity_all_negative <- sensitivity[
  contrast == "SSc-ILD vs IPF",
  all(effect_logit < 0)
]
ssc_ipf_loo_stable_panel_n <- loo[
  contrast == "SSc-ILD vs IPF",
  .(no_flip_in_both_cohorts = all(direction_flip_n == 0)),
  by = panel
][no_flip_in_both_cohorts == TRUE, .N]
ssc_ipf_loo_flip_rows <- loo[
  contrast == "SSc-ILD vs IPF" & direction_flip_n > 0
]
stopifnot(
  ci_cross_zero_n == 18L,
  fdr_n == 0L,
  ssc_ipf_primary_all_negative,
  ssc_ipf_sensitivity_all_negative,
  ssc_ipf_loo_stable_panel_n == 5L,
  all(ssc_ipf_loo_flip_rows$panel == "IgA_mucosal_plasma")
)

fmt <- function(x, digits = 3) {
  ifelse(is.na(x), "NA", formatC(x, digits = digits, format = "f"))
}
table_html <- function(x, caption = NULL, digits = 4) {
  html <- kable(x, format = "html", caption = caption, digits = digits, escape = TRUE,
                table.attr = 'class="table table-striped table-hover table-sm"')
  knitr::asis_output(
    paste0('<div style="width:100%;max-height:480px;overflow:auto;">', html, '</div>')
  )
}
image_tag <- function(name, alt) {
  src <- knitr::image_uri(file.path(figure_dir, name))
  sprintf('<img src="%s" alt="%s" style="width:100%%;height:auto;"/>', src, alt)
}
```

::: {.callout-note}
## FINAL — manually approved

This exploratory patient-level comparison and its statistical interpretation were approved after manual review. The report reuses immutable frozen patient-level outputs and does **not** recalculate Bin-level scores, alter panels or thresholds, call TLS, or add samples/tests.
:::

## Executive audit summary

The primary cohort contains 15 prespecified patients (HC = 4, IPF = 4, SSc-ILD = 7) after removing the five independently flagged low-quality samples and the independently identified extreme-depth sample `IPF/FO22-1-09404`. The J2/L3-only sensitivity cohort contains 13 patients (HC = 3, IPF = 4, SSc-ILD = 6) after additionally removing the two remaining K8 samples, `HC/NL-50` and `SSC/24-1-18170A2`.

All six frozen panels and their existing strict same-bin/rook definitions were reused. The primary endpoint was transformed exactly as preregistered:

`logit((strict_same_bin_supported_bin_n + 0.5) / (total_bin_n + 1))`.

Each panel used the patient-level model `transformed_endpoint ~ disease_group + log(sample_total_counts) + chip_id`, with HC reference and HC3 robust standard errors. The 18 primary P values were adjusted once as one BH family.

**All `r ci_cross_zero_n`/18 primary 95% confidence intervals crossed zero, and `r fdr_n`/18 tests reached BH-FDR < 0.05.** No nominal P value is interpreted alone as a reliable finding.

For **SSc-ILD vs IPF**, the adjusted effect was negative in all six panels in both the Primary15 and J2/L3-only13 complete-cohort models. Leave-one-patient-out refits preserved that direction for five of six panels; `IgA_mucosal_plasma` was the exception, with three direction-flipping refits in each cohort. The cross-panel negative pattern is therefore an exploratory directional trend only and is not uniformly LOO-stable: it **cannot be interpreted as an established disease difference**.

No further adjustment of the model, sample set, markers or thresholds will be made in response to these results.

Panels with all three adjusted contrast directions preserved after excluding K8 were: **`r paste(all_consistent, collapse = ", ")`**. Direction flips occurred only for `r if (nrow(flipped)) paste(paste0(flipped$panel, " (", flipped$contrast, ")"), collapse = "; ") else "none"`; these estimates were close to zero and therefore especially sensitive to modest technical adjustment changes.

## Frozen inputs and cohorts

```{r}
table_html(read_table("00_data_source_and_frozen_rules_manifest.tsv"), "Frozen analysis rules")
```

```{r}
table_html(read_table("02_primary15_sample_list.tsv"), "Primary cohort: 15 patients")
```

```{r}
table_html(read_table("03_j2_l3_only13_sample_list.tsv"), "J2/L3-only sensitivity cohort: 13 patients")
```

```{r}
table_html(read_table("01_frozen_six_panel_marker_manifest.tsv"), "Frozen panel marker manifest")
```

## Patient-level descriptive data

All points below are patients; there are no Bin-level pseudo-replicate error bars. Zero same-bin support is retained and receives only the prespecified 0.5 continuity correction for modeling. The fixed descriptive sparse flag is `supported_bin_n <= 1` and is not used for exclusion.

```{r, results='asis'}
cat(image_tag("01_primary15_patient_scatter.png", "Primary15 patient-level scatter"))
```

```{r, results='asis'}
cat(image_tag("02_j2_l3_only13_patient_scatter.png", "J2/L3-only13 patient-level scatter"))
```

```{r}
table_html(
  group_stats,
  "Group descriptive statistics: median, IQR, minimum, maximum and zero count"
)
```

```{r}
table_html(
  patients[, .(
    cohort, sample_id, group, chip_id, sample_total_counts, panel,
    strict_same_bin_supported_bin_n, total_bin_n, strict_same_bin_fraction,
    strict_rook_fraction, panel_cpm_total, same_bin_zero,
    extremely_sparse_predefined
  )],
  "Complete patient-level endpoints"
)
```

## Primary15 adjusted disease contrasts

`Mature_B_core` is the primary biological endpoint. The other five panels are secondary exploratory endpoints. Effects and confidence intervals are on the corrected-logit scale. Adjusted fraction differences are standardized over the observed cohort technical covariates and are included only as an aid to interpretation.

```{r, results='asis'}
cat(image_tag("03_primary15_forest_18_contrasts.png", "Primary15 forest plot"))
```

```{r}
primary_display <- primary[, .(
  panel, endpoint_role, contrast,
  effect_logit, ci_low, ci_high, p_value, bh_fdr_18,
  adjusted_fraction_numerator, adjusted_fraction_denominator,
  adjusted_fraction_difference, fraction_difference_ci_low,
  fraction_difference_ci_high
)]
table_html(primary_display, "Primary15 model: all 18 prespecified contrasts")
```

```{r}
table_html(predictions, "Adjusted group predictions / marginal means")
```

## J2/L3-only sensitivity and K8 assessment

The 13-patient model is unchanged except that K8 is absent and `chip_id` therefore contains only J2 and L3. Nominal P values are shown descriptively; they are not used for a new significance screen.

```{r, results='asis'}
cat(image_tag("04_primary15_vs_j2_l3_only13_effect_direction.png", "Primary15 versus J2/L3-only13 effects"))
```

```{r}
table_html(stability, "Primary15 versus J2/L3-only13 effect-direction stability")
```

```{r}
table_html(sensitivity, "J2/L3-only13 sensitivity model results")
```

The K8 sensitivity is not uniformly null: four panels retain all three directions, whereas `Mature_B_core` (`SSc-ILD vs HC`) and `IgA_mucosal_plasma` (`IPF vs HC`) flip direction. Both flipped primary15 estimates are small relative to their confidence intervals. Thus those particular comparisons are K8-sensitive, while the remaining directions are not dependent on K8 inclusion by this check. Direction preservation does not imply biological validation.

## Same-bin, rook and CPM descriptive direction checks

Rook support and panel CPM are descriptive sensitivities only. No additional P values were generated.

```{r, results='asis'}
cat(image_tag("05_same_bin_rook_cpm_direction_consistency.png", "Same-bin rook CPM directions"))
```

```{r}
table_html(direction, "Disease-group median directions for same-bin, rook and CPM")
```

## Diagnostics and patient influence

All 12 full models were full-rank and yielded finite HC3 covariance estimates. Influence flags use fixed conventional cutoffs (leverage > 2p/n, Cook distance > 4/n, or |studentized residual| > 2); they trigger review, not deletion.

```{r, results='asis'}
cat(image_tag("06_model_influence_diagnostics.png", "Patient influence diagnostics"))
```

```{r}
table_html(diagnostics, "Model diagnostics")
```

```{r}
table_html(influence[flagged_any == TRUE], "Patients flagged by at least one influence diagnostic")
```

```{r}
table_html(loo, "Leave-one-patient-out effect stability")
```

Several near-zero contrasts change direction under leave-one-patient-out refitting, particularly IPF-vs-HC estimates for IgA, GC and ILD-plasma panels. The most recurrent high-influence labels include `SSC/05957/17B`, `SSC/23-105334B1` and `IPF/FO23-1-09473`, but no patient was removed. Two primary15 leave-one-out refits per contrast are rank-deficient because dropping either of the two remaining K8 patients leaves the K8 chip represented by only one disease group; these failed refits are explicitly retained as unstable rather than silently omitted. Overall, the wide confidence intervals, diagnostic flags and leave-one-out changes argue against a claim that any observed contrast is robustly established.

## Interpretation boundaries

- Marker co-capture fraction is a technical/descriptive endpoint and is not the physical area occupied by B cells.
- Plasma and IgA panels may be dominated by abundant transcripts such as `JCHAIN` and `IGHA1`.
- `GC_activated_B` cannot be interpreted directly as a germinal center or TLS.
- K8 previously showed systematically higher capture; the J2/L3-only13 analysis is therefore a key sensitivity analysis, not a replacement cohort selected by outcome.
- Sample sizes are small. Effect size, 95% CI, patient scatter, influence diagnostics and 15/13 direction stability take precedence over nominal P values.
- Only effects that are direction-stable, not driven by a single patient and persistent after technical adjustment would be eligible for a later spatial review. This DRAFT does not identify such effects as validated biology.

## Output validation

```{r}
table_html(validation, "Automated output checks")
```

This is the approved final local report. The corresponding ELN document remains a draft pending its separate approval gate; no remote ELN entry or Git push was created.

```{bash render}
#| eval: false
# Run from the project scripts directory. This renders the report only and
# reads existing TSV/PNG outputs; it does not execute the analysis model.
$env:BLINEAGE_DISEASE_RESULT_DIR = (Resolve-Path "..\results\reanalysis\bin50_cleaned_blineage_patient_disease_comparison_260823192715").Path
quarto render 394_cleaned_blineage_patient_disease_comparison_report.qmd --output FINAL_394_cleaned_blineage_patient_disease_comparison_report_260823192715.html
```
