---
title: "L3 K=8 Spatial Niche Architecture"
author: "Codex"
date: today
format:
  html:
    toc: true
    toc-float: true
    df-print: paged
    code-fold: true
    theme: bootstrap
    embed-resources: true
engine: knitr
execute:
  echo: true
---

<!--
Render command (PowerShell; set the timestamp immediately before rendering):
quarto render scripts/1012_l3_k8_spatial_architecture_report.qmd --output-dir results
Rename the generated HTML to results/L3_K8_SPATIAL_ARCHITECTURE_REPORT_<YYMMDDHHMMSS>.html.
-->

```{r setup}
project_root <- normalizePath("..", mustWork = TRUE)
candidate_dirs <- Sys.glob(file.path(
  project_root,
  "results",
  "reanalysis",
  "L3_UNBIASED_NICHE_RESOLUTION_20261007_085730",
  "K08_SPATIAL_ARCHITECTURE_*"
))
candidate_dirs <- candidate_dirs[file.exists(file.path(candidate_dirs, "COMPLETE.ok"))]
stopifnot(length(candidate_dirs) > 0)
result_dir <- tail(sort(candidate_dirs), 1)

read_tsv <- function(filename) {
  read.delim(
    file.path(result_dir, filename),
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
}

continuity <- read_tsv("01_K08_NICHE_SPATIAL_CONTINUITY.tsv")
cross_patient <- read_tsv("05_K08_CROSS_PATIENT_ADJACENCY_SUMMARY.tsv")
n3_summary <- read_tsv("06_K08_N3_NEIGHBOR_SUMMARY.tsv")
priority <- read_tsv("07A_K08_GLOBAL_SPATIAL_PRIORITY.tsv")
audit <- read_tsv("09_K08_SPATIAL_ARCHITECTURE_FINAL_AUDIT.tsv")
```

## Scope and provenance

This report summarizes the completed L3 K=8 spatial architecture analysis. It reuses the frozen K=8 assignment, 39,982 neighborhood-evaluable Bin50, and physical `x_um`/`y_um` coordinates. Frozen aggregate labels and boundaries were not read.

Direct Queen adjacency was defined only within the same patient and ROI. A pair of distinct Bin50 was adjacent when `abs(dx) <= 25 um` and `abs(dy) <= 25 um`. Each physical edge was counted once. The patient/ROI-stratified null used 999 label permutations while preserving niche counts.

## Spatial continuity

```{r continuity-summary}
continuity_summary <- aggregate(
  cbind(largest_component_fraction, fragmentation_index, median_component_size) ~ niche,
  continuity,
  median
)
continuity_summary <- continuity_summary[
  order(-continuity_summary$largest_component_fraction),
]
knitr::kable(continuity_summary, digits = 3)
```

## All 28 heterotypic niche pairs

```{r global-priority}
priority_display <- priority[, c(
  "global_priority_rank",
  "niche_pair",
  "median_log2_enrichment",
  "positive_patient_n",
  "negative_patient_n",
  "consistent_direction_patient_n",
  "single_patient_driven",
  "both_endpoints_umi_robust",
  "priority_interpretation"
)]
knitr::kable(priority_display, digits = 3)
```

The exchangeable-label null breaks niche-label spatial autocorrelation and does not preserve the overlap among adjacent 50 um input neighborhoods. Therefore, observed/expected adjacency describes departure from label exchangeability; it is not independent validation of a biological interface or cell-cell interaction.

## N3 targeted follow-up

```{r n3-summary}
n3_display <- n3_summary[, c(
  "neighboring_niche",
  "all_l3_unique_direct_edge_count",
  "median_log2_enrichment",
  "positive_patient_n",
  "negative_patient_n",
  "direction_consistency",
  "single_patient_driven"
)]
knitr::kable(n3_display, digits = 3)
```

## Validation

```{r final-audit}
knitr::kable(audit)
stopifnot(all(audit$passed))
```

## Output files

The full result directory is:

```{r result-directory}
result_dir
list.files(result_dir, recursive = TRUE)
```

## Session information

```{r session-info}
sessionInfo()
```
