---
title: "J2 Unbiased 50 um Neighborhood Niche Resolution"
author: "Codex"
date: today
format:
  html:
    toc: true
    toc-float: true
    df-print: paged
    code-fold: true
    theme: bootstrap
    embed-resources: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

<!--
Render command (PowerShell; set the timestamp immediately before rendering):
& "D:/bb/RSTUDIO/resources/app/bin/quarto/bin/quarto.exe" render scripts/1017_j2_unbiased_niche_resolution_report.qmd --output-dir results --no-clean
Rename the generated HTML to results/J2_UNBIASED_NICHE_RESOLUTION_REPORT_<YYMMDDHHMMSS>.html.
-->

```{r setup}
library(data.table)
library(knitr)

script_dir <- dirname(knitr::current_input(dir = TRUE))
project_dir <- normalizePath(file.path(script_dir, ".."), mustWork = TRUE)
result_dir <- file.path(
  project_dir,
  "results/reanalysis/J2_UNBIASED_NICHE_RESOLUTION_20261008_110522"
)

input_audit <- fread(file.path(result_dir, "00_J2_INPUT_AUDIT.tsv"))
neighborhood_audit <- fread(file.path(result_dir, "01_J2_NEIGHBORHOOD_AUDIT.tsv"))
recommendation <- fread(file.path(result_dir, "04_J2_AUDITED_K_RECOMMENDATION.tsv"))
annotations <- fread(file.path(result_dir, "11_J2_PRIMARYK_CONSERVATIVE_ANNOTATIONS.tsv"))
cross_patient <- fread(file.path(result_dir, "09_J2_PRIMARYK_CROSS_PATIENT_ENRICHMENT.tsv"))
final_audit <- fread(file.path(result_dir, "13_J2_FINAL_AUDIT.tsv"))

primary <- recommendation[role == "primary"][1]
lower <- recommendation[role == "lower-resolution comparator"][1]
higher <- recommendation[role == "higher-resolution sensitivity"][1]

audit_value <- function(item_name) {
  final_audit[item == item_name, value][[1]]
}
```

## Scope and provenance

This report documents the completed, chip-specific discovery analysis for
Y40105J2. It used frozen Reference18 continuous weights for Bin50 with UMI >= 50,
the official 50 um radius neighborhood definition, native VoltRon CLR features,
and independent K-means fits for K=3--12. The center Bin was excluded from its own
neighborhood, matching the audited L3 framework.

- UMI-eligible Bin50: `r format(as.integer(audit_value("total_umi50_bin_n")), big.mark = ",")`
- Neighborhood-evaluable Bin50: `r format(as.integer(audit_value("evaluable_bin_n")), big.mark = ",")`
- Patients: `r audit_value("patient_n")` (`r audit_value("disease_counts")`)
- Reference18 states: 18
- Radius: `r audit_value("neighborhood_radius_um")` um
- Random seeds per K: `r audit_value("seeds_per_K")`

Frozen lymphoid aggregate labels were not read. Disease labels were retained only
as metadata and were not used for neighborhood construction, clustering, K
selection, or niche annotation. This analysis did not perform disease-significance
testing, DEG, GSEA, pathway analysis, or cell-cell communication analysis.

```{r input_audit_table}
kable(
  input_audit[, .(
    patient_id,
    disease,
    total_umi50_bin_n,
    evaluable_radius50_bin_n,
    roi_n,
    missing_coordinate_n,
    missing_reference18_weight_n
  )],
  caption = "J2 patient-level input audit"
)
```

```{r neighborhood_audit_table}
kable(neighborhood_audit, caption = "Audited 50 um neighborhood definition")
```

## K=3--12 resolution audit

```{r k_resolution_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/01_J2_K_RESOLUTION_AUDIT_EN.png"
))
```

```{r recommendation_table}
kable(
  recommendation[, .(
    k,
    role,
    converged_seed_n,
    seed_fit_n,
    repeat_ari_median,
    repeat_nmi_median,
    betweenss_fraction,
    min_niche_fraction,
    min_patient_coverage_n,
    below_one_percent_niche_n,
    single_patient_niche_n,
    within_patient_umi_eta_squared,
    audited_reason
  )],
  digits = 3,
  caption = "Audited J2 resolution recommendation and comparators"
)
```

K=`r primary$k` is the audited primary resolution. It converged in
`r primary$converged_seed_n`/`r primary$seed_fit_n` fits, had median pairwise ARI
`r sprintf("%.3f", primary$repeat_ari_median)`, covered all seven patients in every
niche, and had a smallest niche fraction of
`r sprintf("%.1f%%", 100 * primary$min_niche_fraction)`. K=`r lower$k` is retained
as an under-resolution comparator; K=`r higher$k` is retained only as a
higher-resolution sensitivity analysis. Disease separation was not a K-selection
criterion.

## Primary-K Reference18 composition

```{r composition_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/02_J2_PRIMARYK_REFERENCE18_COMPOSITION_EN.png"
))
```

```{r annotation_table}
kable(
  annotations[, .(
    niche,
    proposed_annotation,
    core_positive_states,
    positive_patient_consistency,
    umi_robust,
    any_state_single_patient_driven,
    core_state_single_patient_driven,
    confidence,
    interpretation_limit
  )],
  caption = "Conservative primary-K niche annotations"
)
```

The five niches represent mixed neighborhood-composition gradients rather than
pure cell types. The most reproducible axes include AT1 enrichment in N1, B
enrichment in N2, lymphatic-endothelial enrichment in N4, and blood-endothelial
enrichment in N5. N3 has a broader mixed airway/stromal/immune profile. No core
state was classified as single-patient-driven.

## Cross-patient consistency

```{r patient_enrichment_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/03_J2_PRIMARYK_PATIENT_WISE_ENRICHMENT_EN.png"
))
```

```{r direction_consistency_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/04_J2_PRIMARYK_CROSS_PATIENT_DIRECTION_CONSISTENCY_EN.png"
))
```

```{r core_consistency_table}
core_states <- unique(unlist(strsplit(
  paste(annotations$core_positive_states, collapse = ";"),
  ";\\s*"
)))
core_summary <- cross_patient[
  reference18_state %in% core_states & positive_patient_n >= 6,
  .(
    niche,
    reference18_state,
    median_enrichment,
    positive_patient_n,
    evaluable_patient_count,
    loo_direction_stable,
    single_patient_driven
  )
]
kable(
  core_summary[order(niche, -median_enrichment)],
  digits = 4,
  caption = "Cross-patient consistency for recurrent core-state enrichments"
)
```

`any_state_single_patient_driven` and `core_state_single_patient_driven` are
reported separately so that a weak peripheral state cannot invalidate a stable
core composition axis.

## UMI sensitivity

```{r umi_sensitivity_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/05_J2_PRIMARYK_UMI_SENSITIVITY_EN.png"
))
```

The core directions were retained after adjustment for scaled log1p(Bin UMI) and
within-patient UMI-quantile stratification. These checks assess sensitivity; they
do not establish that UMI effects have been completely removed.

## Interpretation boundaries

J2 independently recapitulated broad B-enriched, alveolar-associated,
endothelial-associated, airway-associated, and stromal/immune mixed themes. This
is a blind descriptive comparison with L3, not a formal cross-chip alignment.
Niche numbers are chip-specific and cannot be mapped across chips by number.

The analysis supports proceeding to a separate J2 spatial architecture audit and
then an independently fitted Y40102K8 analysis. Cross-chip consensus and biological
pathway analyses remain outside this report.

## Output validation

```{r final_audit_table}
kable(final_audit, caption = "Final J2 output audit")
```

All saved audit rows passed. The analysis directory also contains seed-level fit
diagnostics, all-K assignments, patient-specific composition tables, and complete
session information.

## Session information

```{r session_info}
sessionInfo()
```
