---
title: "J2 K=5 Spatial Niche Architecture Audit"
author: "Codex"
date: today
format:
  html:
    toc: true
    toc-float: true
    df-print: paged
    code-fold: true
    theme: bootstrap
    embed-resources: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

<!--
Render command (PowerShell; set the timestamp immediately before rendering):
& "D:/bb/RSTUDIO/resources/app/bin/quarto/bin/quarto.exe" render scripts/1019_j2_k5_spatial_architecture_report.qmd --output-dir results --no-clean
Rename the generated HTML to results/J2_K5_SPATIAL_ARCHITECTURE_REPORT_<YYMMDDHHMMSS>.html.
-->

```{r setup}
library(data.table)
library(knitr)

script_dir <- dirname(knitr::current_input(dir = TRUE))
project_dir <- normalizePath(file.path(script_dir, ".."), mustWork = TRUE)
result_dir <- file.path(
  project_dir,
  "results/reanalysis/J2_UNBIASED_NICHE_RESOLUTION_20261008_110522",
  "K05_SPATIAL_ARCHITECTURE_20261008_120148"
)

continuity <- fread(file.path(result_dir, "01_K05_NICHE_SPATIAL_CONTINUITY.tsv"))
adjacency_counts <- fread(file.path(result_dir, "02_K05_NICHE_ADJACENCY_COUNTS.tsv"))
patient_enrichment <- fread(file.path(
  result_dir,
  "04_K05_PATIENT_NICHE_ADJACENCY_ENRICHMENT.tsv"
))
cross_patient <- fread(file.path(
  result_dir,
  "05_K05_CROSS_PATIENT_ADJACENCY_SUMMARY.tsv"
))
continuity_summary <- fread(file.path(
  result_dir,
  "06_K05_CROSS_PATIENT_CONTINUITY_SUMMARY.tsv"
))
audit <- fread(file.path(
  result_dir,
  "08_K05_SPATIAL_ARCHITECTURE_FINAL_AUDIT.tsv"
))
evaluable_bin_n <- continuity[, unique(patient_evaluable_bin_n), by = patient_id][, sum(V1)]
unique_edge_n <- adjacency_counts[scope == "all_J2", unique(total_unique_edge_n)]
```

## Scope and provenance

This audit reuses the frozen Y40105J2 K=5 assignment. It does not rerun
clustering or RCTD. The analysis includes all five niches symmetrically and all
10 heterotypic niche pairs.

- Neighborhood-evaluable Bin50: `r format(evaluable_bin_n, big.mark = ",")`
- Patients: `r uniqueN(continuity$patient_id)`
- Patient/ROI-specific Queen edges: `r format(unique_edge_n, big.mark = ",")`
- Direct adjacency: same patient and ROI, `abs(dx) <= 25 um` and `abs(dy) <= 25 um`, excluding self-pairs
- Null: 999 niche-label permutations within each patient and ROI

Frozen aggregate labels and boundaries were not read. Disease labels were used
only for patient labels in figures. No disease-significance, DEG/GSEA, pathway,
CellChat, ligand-receptor, or cross-chip consensus analysis was performed.

## Spatial niche maps

The complete seven-page map is stored at
`figures/01_K05_ALL_PATIENT_SPATIAL_NICHE_MAPS_EN.pdf`. Each page uses the same
K=5 color key, actual Bin50 coordinates, and a 25 x 25 um tile size.

## Connected components and fragmentation

```{r continuity_table}
kable(
  continuity_summary,
  digits = 3,
  caption = "Cross-patient median K=5 spatial-continuity metrics"
)
```

```{r component_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/02_K05_NICHE_COMPONENT_SIZE_EN.png"
))
```

```{r largest_component_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/03_K05_NICHE_LARGEST_COMPONENT_FRACTION_EN.png"
))
```

N2 had the highest cross-patient median largest-component fraction, followed by
N4 and N1. Absolute largest-component fractions remained modest, indicating
that every K=5 label occurred in multiple disconnected patches rather than one
continuous tissue compartment.

## Direct niche adjacency

```{r adjacency_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/04_K05_NICHE_ADJACENCY_HEATMAP_EN.png"
))
```

The diagonal captures strong homotypic adjacency. Because K=5 labels derive
from overlapping 50 um neighborhood-composition profiles, high self-adjacency
is expected and is not treated as independent biological validation.

## Cross-patient spatial consistency

```{r cross_patient_table}
kable(
  cross_patient[, .(
    niche_pair,
    median_log2_enrichment,
    positive_patient_n,
    negative_patient_n,
    evaluable_patient_n,
    direction_consistency,
    max_absolute_share,
    single_patient_driven
  )],
  digits = 3,
  caption = "All 10 heterotypic K=5 niche-pair results"
)
```

```{r cross_patient_figure, echo=FALSE, out.width="100%"}
knitr::include_graphics(file.path(
  result_dir,
  "figures/05_K05_CROSS_PATIENT_ADJACENCY_ENRICHMENT_EN.png"
))
```

Nine of 10 heterotypic pairs retained a negative observed/expected direction in
at least six of seven patients. No heterotypic pair showed recurrent positive
enrichment. N1--N3 had the largest negative median displacement. These results
describe spatial separation relative to an exchangeable-label null; they do not
demonstrate cell-cell communication or causal exclusion.

## Validation and limitations

```{r validation_table}
kable(audit, caption = "Final spatial-architecture output audit")
```

All 30 audit rows passed. The permutation checkpoint contains 105 patient-pair
rows: seven patients multiplied by 15 total pair types, including 70 patient by
heterotypic-pair observations. The initial run completed all seven permutation
blocks and stopped during a downstream sort; the corrected continuation reused
the saved complete permutation table and did not repeat the null calculation.

The label-permutation null disrupts endogenous spatial autocorrelation and does
not preserve the overlap among adjacent 50 um input neighborhoods. Consequently,
the audit supports organized niche-label fields and interface preferences, not
independent validation of cell identities or interactions.

## Session information

```{r session_info}
sessionInfo()
```
