---
title: "CDR3 per-Bin Top20 heatmaps"
author: "Codex"
date: today
format:
  html:
    toc: true
    toc-float: true
    df-print: paged
    code-fold: true
    theme: bootstrap
    embed-resources: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

<!--
Render with the timestamping wrapper:
powershell.exe -NoProfile -ExecutionPolicy Bypass -File scripts/630_render_cdr3_per_bin_top20_heatmap_report.ps1
-->

## Scope

This report documents the one-off generation and validation of recurrent and
full heatmaps from the frozen per-Bin Top20 gene tables. It does not read a raw
expression matrix, rank genes, or recompute upstream IGH/CDR3 results.

```{r setup}
project_root <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
  winslash = "/",
  mustWork = TRUE
)
output_dir <- file.path(
  project_root,
  "results/reanalysis/CDR3_PER_BIN_TOP20_HEATMAP_FINAL_20260912_141500"
)

required_files <- file.path(
  output_dir,
  c(
    "TOP20_HEATMAP_QC.tsv",
    "TOP20_HEATMAP_QC_SUMMARY.txt",
    "RUN_LOG.txt",
    "MANIFEST.tsv",
    "run_config.json"
  )
)
stopifnot(all(file.exists(required_files)))

qc <- read.delim(
  file.path(output_dir, "TOP20_HEATMAP_QC.tsv"),
  sep = "\t",
  check.names = FALSE
)
manifest <- read.delim(
  file.path(output_dir, "MANIFEST.tsv"),
  sep = "\t",
  check.names = FALSE
)
run_log <- readLines(file.path(output_dir, "RUN_LOG.txt"), warn = FALSE)
qc_summary <- readLines(
  file.path(output_dir, "TOP20_HEATMAP_QC_SUMMARY.txt"),
  warn = FALSE
)

stopifnot(nrow(qc) > 0L, all(qc$passed), nrow(manifest) > 0L)
```

## Frozen definitions

```{r frozen-summary, results='asis'}
cat(paste0("- ", qc_summary, collapse = "\n"))
```

The heatmap value is `log1p(raw_count)` for genes already present in each
physical Bin50's frozen Top20 list. Blank cells indicate that a gene was not
ranked in that Bin50's Top20; they do not establish nondetection. A recurrent
gene appears in the Top20 of at least two distinct physical Bin50s, keyed by
`sample_id + Bin50_id`. Columns and rows are not clustered, and no gene-wise
z-score is applied.

## Main recurrent heatmaps

```{r recurrent-figures, echo=FALSE, out.width='100%', fig.align='center'}
recurrent_paths <- file.path(
  output_dir,
  c(
    "FIG1_ALL69_RECURRENT_TOP20_HEATMAP.png",
    "FIG2_SHARED7_RECURRENT_TOP20_HEATMAP.png",
    "FIG3_REPEATED11_RECURRENT_TOP20_HEATMAP.png"
  )
)
stopifnot(all(file.exists(recurrent_paths)))
knitr::include_graphics(recurrent_paths)
```

## Supplementary full heatmaps

```{r full-figures, echo=FALSE, out.width='100%', fig.align='center'}
full_paths <- file.path(
  output_dir,
  c(
    "SUPP_FIG1_ALL69_FULL_TOP20_HEATMAP.png",
    "SUPP_FIG2_SHARED7_FULL_TOP20_HEATMAP.png",
    "SUPP_FIG3_REPEATED11_FULL_TOP20_HEATMAP.png"
  )
)
stopifnot(all(file.exists(full_paths)))
knitr::include_graphics(full_paths)
```

## Quality control

```{r qc-table}
qc
```

```{r qc-assertions}
data.frame(
  qc_rows = nrow(qc),
  qc_failures = sum(!qc$passed),
  manifest_rows = nrow(manifest),
  recurrent_png = sum(grepl("RECURRENT.*[.]png$", manifest$path)),
  full_png = sum(grepl("FULL.*[.]png$", manifest$path)),
  stringsAsFactors = FALSE
)
```

All recorded QC checks passed. The final independent verification also
recomputed every MANIFEST file size and SHA-256 digest and found no mismatch.
All six PNG files decoded successfully, and visual review of the primary All69
recurrent heatmap and longest All69 full heatmap found no truncated or blank
rendering.

## Reproducibility

The exact plotting command was:

```text
D:/bb/R/R-4.5.2/bin/x64/Rscript.exe scripts/628_cdr3_per_bin_top20_heatmaps.R results/reanalysis/CDR3_PER_BIN_GENE_LIST_20260911_161928 results/reanalysis/CDR3_PER_BIN_TOP10_FINAL_20260911_165003 results/reanalysis/CDR3_PER_BIN_TOP10_HEATMAPS_20260911_172841 results/reanalysis/CDR3_PER_BIN_TOP20_HEATMAP_FINAL_20260912_141500
```

```{r run-log, echo=FALSE, results='asis'}
cat("```text\n", paste(run_log, collapse = "\n"), "\n```\n", sep = "")
```

```{r manifest}
manifest
```

## Session information

```{r session-info}
sessionInfo()
```
